---
title: "Body Measures Demo"
output: 
  html_document:
    df_print: kable
    code_folding: hide
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

Let's explore the relationship between body measurements. Are they normally distributed? Bivariate normally distributed? 

The data are from the NHANES last pre-covid measurements in 2017-2018 (see [here](https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/BMX_J.htm) for more info).

# Setup

```{r}
library(tidyverse)
library(shiny)
library(DT)
library(ggplot2)
library(mvtnorm)
```

# Data

```{r}
bm <- read_csv("https://mshsu.github.io/probasets/data/BodyMeasures.csv")
```

```{r}
# Show all data
renderDataTable({
  datatable(bm, options = list(scrollX = TRUE))
})
```

# Variables

```{r}
# List variable names we want to see bvn relationships between
# To be used in reactive dropdown
varnames <- c('BMXWT', 'BMXRECUM', 'BMXHEAD',
              'BMXHT', 'BMXBMI', 'BMXLEG',
              'BMXARML', 'BMXARMC', 'BMXWAIST',
              'BMXHIP')
```

```{r}
# Use list of variable names to choose first variable
selectInput("var1", 
            label = "Select your first variable:",
            choices = varnames, selected = "BMXWT")
```

```{r}
# Use list of variable names to choose second variable
selectInput("var2", 
            label = "Select your second variable:",
            choices = varnames, selected = "BMXBMI")
```

```{r}
# Show all observations with both variables not null
renderDataTable({
  bm %>%
    select(SEQN, input$var1, input$var2) %>%
    drop_na()
})
```

```{r}
# Same as table above, but with only variables
bm2 <- reactive({
  bm %>%
    select(input$var1, input$var2) %>%
    drop_na()
})

# Calculate mean of variable 1, to be used in visualizations
mean1 <- reactive({
  mean(bm2()[[input$var1]], na.rm=TRUE)
})
# Calculate mean of variable 2, to be used in visualizations
mean2 <- reactive({
  mean(bm2()[[input$var2]], na.rm=TRUE)
})
# Calculate sd of variable 1, to be used in visualizations
sd1 <- reactive({
  sd(bm2()[[input$var1]], na.rm=TRUE)
})
# Calculate sd of variable 1, to be used in visualizations
sd2 <- reactive({
  sd(bm2()[[input$var2]], na.rm=TRUE)
})
# Calculate cov, to be used in visualizations
cv <- reactive({
  cov(bm2())[1,2]
})
# Calculate corr, to be used in visualizations
r <- reactive({
  cor(bm2())[1,2]
})
```


# Descriptive Statistics

Mean:

```{r}
# Here we are simply printing out descriptive statistics
renderPrint({
  print(apply(bm2(), 2, mean))
})
```

Standard Deviation:

```{r}
renderPrint({
  print(apply(bm2(), 2, sd))
})
```

Covariance Matrix:

```{r}
renderPrint({
  print(cov(bm2()))
})
```

Correlation Matrix:

```{r}
renderPrint({
  print(cor(bm2()))
})
```

# Marginal Distributions

ORANGE represents the theoretical marginal normal distribution.

BLUE represents the observed data marginal distribution.

```{r}
renderPlot({
  # Here we want to plot the marginal distribution of variable 1
  # We are using the bm2 dataset, which contains all observations with 
  #   both variables selected in the reactive dropdown being not null
  # !!sym(input$var1) changes our chosen variable 1 name from a string
  #   to a literal input that we can use as a variable in aes()
  ggplot(data = bm2(), aes(x = !!sym(input$var1))) + 
    # geom_histogram plots our histogram of the observed data in blue
    geom_histogram(aes(y = ..density..), bins = 10, alpha = 0.5, fill = "blue") +
    # stat_function plots the theoretical marginal distribution
    stat_function(fun = dnorm, 
                  args = list(mean = mean1(), 
                              sd = sd1()), 
                  color = "orange", size = 1.5) +
    # Cosmetic stuff
    labs(title = paste("Histogram of", input$var1), x = input$var1, y = "Density") +
    theme(panel.background = element_rect(fill='transparent'),
          plot.background = element_rect(fill='transparent', color=NA),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank())
})
```

```{r}
renderPlot({
  # We repeat the previous chunk, but for variable 2
  ggplot(data = bm2(), aes(x = !!sym(input$var2))) + 
    geom_histogram(aes(y = after_stat(density)), 
                   bins = 10, alpha = 0.5, fill = "blue") + 
    stat_function(fun = dnorm, 
                  args = list(mean = mean2(), 
                              sd = sd2()), 
                  color = "orange", size = 1.5) +
    labs(title = paste("Histogram of", input$var2), x = input$var2, y = "Density") +
    theme(panel.background = element_rect(fill='transparent'),
          plot.background = element_rect(fill='transparent', color=NA),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank())
})
```

# Joint Distribution

ORANGE represents the theoretical bivariate normal distribution.

BLUE represents the observed data distribution.

GREEN represents the regression line fit to the data.

```{r}
renderPlot({
  validate(
    need(input$var1 != input$var2, "Please select two different variables")
  )
  # mean vector and covariance matrix
  mu <- c(mean1(), mean2())
  sigma <- cov(bm2())
  
  # create grid of x and y values
  x <- seq(mean1()-3*sd1(), mean1()+3*sd1(), length.out = 100)
  y <- seq(mean2()-3*sd2(), mean2()+3*sd2(), length.out = 100)
  xy <- expand.grid(x, y)
  
  # evaluate bivariate normal density
  z <- dmvnorm(xy, mean = mu, sigma = sigma)
  
  # create data frame with x, y, and z values
  df <- data.frame(x = xy[,1], y = xy[,2], z = z)
  
  # create 2D plot
  ggplot() +
    # Plot observed data as a scatterplot
    geom_point(data = bm2(), 
               mapping = aes(!!sym(input$var1), 
                             !!sym(input$var2)),
               color="blue", alpha=0.5
               ) +
    # Plot theoretical bvn distribution as contour plot
    geom_contour(data = df, mapping = aes(x, y, z = z), color="orange", size=1) +
    # Plot regression line
    geom_smooth(data = bm2(), mapping = aes(!!sym(input$var1), 
                                            !!sym(input$var2)),
                method = "lm", formula = y ~ x, 
                color = "green") +
    # Cosmetic stuff
    labs(x = input$var1, y = input$var2) +
    theme(legend.position="none",
          panel.background = element_rect(fill='transparent'),
          plot.background = element_rect(fill='transparent', color=NA),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank())
})
```

# Conditional Distribution (Simulated)

Each of the following graphic elements are generated independent of each other. Both are generated from the theoretical bivariate normal distribution (in ORANGE):

- RED represents the theoretical conditional distribution in the histogram, or its slice on the scatter/contour plot.
- PURPLE represents simulated values.

```{r}
renderUI({
  # We are conditioning variable 1 on variable 2's value
  # Here we choose a value from the range of values in variable 2
  # For computational reasons, we limit this to integer values
  numericInput("condition", paste("Conditional distribution of ", input$var1, 
                                  ", given ", input$var2, " equals:", sep=''),
            min = floor(mean2() - 2*sd2()),
            max = ceiling(mean2() + 2*sd2()),
            value = round(mean2()),
            step = 0.1)
})
```

```{r}
# Calculate conditional mean, variance
cond.mean <- reactive({
  mean1() + sd1() / sd2() * r() * (input$condition - mean2())
})

cond.var <- reactive({
  (1 - r()^2)*sd1()^2
})

# Simulate values from the theoretical bvn distribution we fit to data
sim_vals <- reactive({
   data.frame(rmvnorm(5000000, 
                      mean=c(mean1(), mean2()),
                      sigma=cov(bm2())))
})

# Choose simulated values that match condition
sim_cond <- reactive({
  sim_vals() %>%
    filter(round(X2, digits = 3) == input$condition)
})

sim_samp <- reactive({
  sample_n(sim_vals(), 10000)
})
```

```{r}
renderPlot({
  validate(
    need(input$var1 != input$var2, "Please select two different variables")
  )
  
  # Create a range to plot our theoretical curves on
  x_vals <- seq(round(mean1()-3*sd1(), digits=1), 
                round(mean1()+3*sd1(), digits=1),
                by = 0.1)
  # This dataframe contains values with which to plot the theoretical 
  #   conditional normal distribution, using the values we created above
  #   and the conditional mean/variance
  x_theoretical <- data.frame(x = x_vals) %>%
    mutate(y = dnorm(x, mean=cond.mean(),
                     sd = sqrt(cond.var())))
  
  # Plot our simulated and theoretical distributions
  ggplot() +
    # Plot histogram of simulated observations values for variable 1
    #   that match condition on variable 2
    geom_histogram(data = sim_cond(), map = aes(y=..density.., x=X1),
                   fill = "purple", alpha = 0.5) + 
    # Plot theoretical conditional normal distribution
    geom_line(data = x_theoretical, 
              mapping = aes(x, y), 
              color = "red", size = 1) +
    # Cosmetic stuff
    theme(panel.background = element_rect(fill='transparent'),
          plot.background = element_rect(fill='transparent', color=NA),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank()) +
    xlim(c(mean1()-2.5*sd1(), mean1()+2.5*sd1())) +
    ylim(c(0, max(x_theoretical$y) + 0.01))+
    xlab(input$var1) +
    ylab("Density")
})
```

Conditional Mean:

```{r}
# Print descriptive statistics
renderPrint({
  print(cond.mean())
})
```

Conditional Variance:

```{r}
renderPrint({
  print(cond.var())
})
```


```{r}
renderPlot({
  validate(
    need(input$var1 != input$var2, "Please select two different variables")
  )
  # mean vector and covariance matrix
  mu <- c(mean1(), mean2())
  sigma <- cov(bm2())
  
  # create grid of x and y values
  x <- seq(mean1()-3*sd1(), mean1()+3*sd1(), length.out = 100)
  y <- seq(mean2()-3*sd2(), mean2()+3*sd2(), length.out = 100)
  xy <- expand.grid(x, y)
  
  # evaluate bivariate normal density
  z <- dmvnorm(xy, mean = mu, sigma = sigma)
  
  # create data frame with x, y, and z values
  df <- data.frame(x = xy[,1], y = xy[,2], z = z)
  
  # create 2D plot
  ggplot() +
    # Create scatterplot of simulated values
    geom_point(data = sim_samp(), 
                               mapping = aes(X1, X2), 
                               color="purple", alpha=0.5) +
    geom_contour(data = df, mapping = aes(x, y, z = z), color="orange", size=1) +
    # Create contour plot of theoretical bvn distribution
    geom_hline(yintercept = input$condition, color = "red", size=1) +
    # cosmetic stuff
    labs(x = input$var1, y = input$var2) +
    theme(legend.position="none",
          panel.background = element_rect(fill='transparent'),
          plot.background = element_rect(fill='transparent', color=NA),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank()) +
    xlim(c(mean1()-2.5*sd1(), mean1()+2.5*sd1()))
})
```

